---
title: "01 - Loading data"
subtitle: "R for Bio Data Science"
format: html
author: El Mehdi Baknine (s194533)
date: "2023-11-15"
editor: visual
---

## Load libraries

```{r}
library("tidyverse")
library("here")
library("readxl")
```

## Load data

```{r}
# Load metadata from sheet 1

meta_data < read_excel(
  here("data/_raw/raw_data.xlsx"),
  sheet = "Suppl_Table1",
  skip = 3,
  na = c("NA", "--")
)

# Load RNA-seq expression data from sheet 10
expr_data <- read_excel(
  here("data/_raw/raw_data.xlsx"),
  sheet = "Suppl_Table10",
  skip = 1,
  na = "NA"
)
```

## Combine the metadata and expression sheets from the excel table

First we will need to transpose the RNA-seq gene expression dataframe before joining it with the meta data.

```{r}
# Check if there are duplicated transcripts (refseq)
expr_data |>
  group_by(refseq) |>
  filter(n() > 1) |>
  summarise(n = n()) |>
  arrange(desc(n))
```

We see that there are no duplicated transcript refseq. We are good to go.

```{r}
# Concatenate gene and refseq colums to have one unique transcript ID not to lose information about gene ID
expr_data_conc <- expr_data |>
  unite(gene:refseq, 
   col = "transcript_id", 
   sep = "_",
   remove = TRUE, 
   na.rm = FALSE)
```

```{r}

# We then need to transpose the expression dataframe
# so that is can be joined with the metadata
expr_data_conc_trans <- expr_data_conc |>
  pivot_longer(-transcript_id,
   names_to = "sample_id") |>
  pivot_wider(names_from = transcript_id,
   values_from = value)

# Let's count how many samples have RNA-seq data
meta_data |>
  group_by(`RNA-seq`) |>
  count()

# Before joining we will check which samples will not be
# joined
meta_data |>
  anti_join(expr_data_conc_trans,
   by = join_by(`Sample-ID` == sample_id)) |> 
  filter(`RNA-seq` == "no") |>
  count()
```

We see that the samples that are not possible to join are for samples where the meta data also states that the is no RNA-seq data for the sample. Hence, it makes sense that it is not possible to join these.

```{r}
# Inner join the data
all_data <- meta_data |>
  inner_join(expr_data_conc_trans, 
   by = join_by(`Sample-ID` == sample_id))
```

## Save the combined data

```{r}
# Save in data directory, as .tsv file
write_tsv(all_data,
 here("data/01_dat_load.tsv"))
```
