---
title: "01 - Loading data"
subtitle: "R for Bio Data Science"
format: html
author: Mikkel Niklas Rasmussen (s193518)
date: "2023-11-08"
editor: visual
---

## Load libraries

```{r}
library("tidyverse")
library("here")
library("fs")
library("readxl")
```

## Load data
```{r}
# Load metadata
meta_data <- read_tsv("data/meta_data.tsv")

# Load RNA-seq expression data from sheet 10
expr_data <- read_tsv("data/expr_data.tsv")
```

### Rename columns
We will start by cleaning the column names, as there does not seem to be any sort of pattern in the naming scheme. We will use the `janitor::clean_names()` function for this task.

```{r}
meta_data <- meta_data |>
  janitor::clean_names()
```


## Check of NAs in the dataset
We need to 
```{r}
meta_data |>
  summarise(
    across(everything(), ~ sum(is.na(.)))
  ) |>
  pivot_longer(everything()) |>
  filter(value > 0) |>
  arrange(desc(value))
```

We see that there are quite a few NAs in the meta data table. Specially, the age is missing for one of the samples. However, we might not need to use all of the these columns for analysis and visualizations, so we will leave them here for now. 

Duplictes of ID


### Transpose expression data


### Inner join expression with Metadata

### Clean weird variables


