---
title: "02 - Cleaning data"
subtitle: "R for Bio Data Science"
format: html
author: Mikkel Niklas Rasmussen (s193518)
date: "2023-11-14"
editor: visual
---

## Load libraries

```{r}
library("tidyverse")
library("here")
library("readxl")
source(here("R/99_proj_func.R"))
```

## Load data

```{r}
# Load data
data <- read_tsv(here("data/01_dat_load.tsv"))
```

### Rename columns

We will start by cleaning the column names, as there does not seem to be any sort of pattern in the naming scheme. We will use the `janitor::clean_names()` function for this task, which is used inside of the custom \`clean_selected_names

```{r}
# Cleaning column names
data <- data |> clean_selected_cols(1:29)
```

## Check of NAs in the dataset

```{r}
data |>
  summarise(
    across(everything(), ~ sum(is.na(.)))
  ) |>
  pivot_longer(everything()) |>
  filter(value > 0) |>
  arrange(desc(value))
```

We see that there are quite a few NAs in the meta data table. Specially, the age is missing for one of the samples. However, we might not need to use all of the these columns for analysis and visualizations, so we will leave them here for now. Comment : The missing age is from a sample that is not part of the 81 samples with RNA_seq

## Check whether there are duplictes of the sample ID

```{r}
# Check for duplicated sample IDs
data |>
  group_by(sample_id) |>
  filter(n() > 1)
```

There are no duplicates

## Clean weird variables

```{r}
# Define columns containing paratheses in the value cells
cols_to_clean <- c(
  "snp_array_6_0_affymetrix",
  "rna_seq",
  "whole_genome_sequencing",
  "whole_exome_sequencing",
  "smoking_status"
)

# Remove references in parentheses in the these columns
data <- data |>
  mutate_at(
    cols_to_clean,
    str_remove,
    pattern = "\\s*\\(.*\\)"
  )

data <- data |> 
  mutate(smoking_history_pack_years = case_when(
    smoking_history_pack_years == "10>PY" ~ 10,
    TRUE ~ as.numeric(smoking_history_pack_years)
  )) 
```

## Save cleaned data

```{r}
# Save as .tsv file
write_tsv(
  data,
  here("data/02_dat_clean.tsv")
)
```
