---
title: "06 - Create heatmap analysis"
subtitle: "R for Bio Data Science"
format: html
author: Thomas Tams (s204540)
date: "2023-11-21"
editor: visual
---

## Load libraries

```{r}
library('tidyverse')
library("here")
source(here("R/99_proj_func.R"))
```

## Load data

```{r}
# Load data
data <- read_tsv(here("data/02_dat_clean.tsv"))
```

## Get expression data from all data

```{r}
expression_data <- data |> 
  select(1, 30:ncol(data))
```

## Log transform expression data with log2(1 + FPKM)

```{r}
log_row <- function(row_values, pseudo_count = 0) {
  log2(pseudo_count + row_values)
}

expr_log <- expression_data |>
  group_by(sample_id) |>
  nest() |>
  mutate(data = map(.x = data,
                    .f = ~log_row(row_values = .x,
                                  pseudo_count = 1))) |>
  unnest(data) |>
  ungroup()
```

## Transpose data to have transcript per row

```{r}
expr_log_trans <- expr_log |>
  group_by(sample_id) |>
  pivot_longer(-sample_id,
              names_to = "transcript_id") |>
  pivot_wider(names_from = sample_id,
   values_from = value) |>
  ungroup()
```

## Filter away transcripts with only zeros

```{r}
expr_log_trans_no_zero <- expr_log_trans |>
  filter(!if_all(starts_with("S"), ~ . == 0))
```

## Perform k-means clustering for each transcript expression data

```{r}
# Function to perform k-means clustering on a single row
cluster_row <- function(row, k) {
  row_data <- row
  cluster_result <- kmeans(row_data, centers = k, algorithm = "Lloyd")
  return(list(cluster = cluster_result$cluster, centers = cluster_result$centers))
}

# Clustering the expression levels for each transcript
# into two groups (low and high expression)
cluster_data <- expr_log_trans_no_zero |>
  group_by(transcript_id) |>
  nest() |>
  ungroup() |>
  mutate(cluster_data = map(.x = data,
                           ~cluster_row(unlist(.x), k=2)))

cluster_tidy <- cluster_data |>
  mutate(cluster = map(cluster_data, "cluster"),
         centers = map(cluster_data, "centers")) |>
  unnest_wider(cluster) |>
  unnest_wider(centers) |>
  select(-c(data, cluster_data)) |>
  rename(cluster_1 = '1',
         cluster_2 = '2')
```

## Find which transcripts have at least 6 expression values in each cluster

```{r}
# Select transcripts with a state fold change of > 3
fold_change <- function(val_1, val_2){
  fold_change <- max(val_1, val_2) - min(val_1, val_2)
  return(fold_change)
}

above_state_fold <- cluster_tidy |>
  mutate(fold_change = map2(.x = cluster_1,
                            .y = cluster_2,
                            .f = ~fold_change(val_1 = .x,
                                              val_2 = .y))) |>
  filter(!fold_change == Inf) |>
  filter(fold_change > 3)

# Pivot longer for grouping by transcripts and cluster groupings
above_state_fold_long <- above_state_fold |>
  pivot_longer(cols = starts_with("S"),
               names_to = "sample_id",
               values_to = "cluster_int")


# Count number of expression values in each cluster
# for all transcripts and select transcript with with
# more than 6 in each group
selected_transcripts <- above_state_fold_long |>
  group_by(transcript_id) |>
  summarise(n_c1 = sum(case_when(cluster_int == 1 ~ 1,
                                TRUE ~ 0)),
            n_c2 = sum(case_when(cluster_int == 2 ~ 1,
                                 TRUE ~ 0))) |> 
  filter(n_c1 >= 6 & n_c2 >= 6) |>
  pull(transcript_id)
```

## Save selected transcripts expression data to csv

```{r}
selected_transcript_expr <- expr_log_trans |>
  filter(transcript_id %in% selected_transcripts)

selected_transcript_expr |>
  write_tsv(here("data/06_selected_transcript_expr.tsv"))
```

## Load expression if already exists

```{r}
if(!exists('selected_transcript_expr')){
  selected_transcript_expr <- 
    read_tsv(here("data/06_selected_transcript_expr.tsv"))
}
```

```{r}
expr_matrix <- as.matrix(selected_transcript_expr[, -1])
rownames(expr_matrix) <- selected_transcript_expr$transcript_id

expr_matrix |>
  pheatmap::pheatmap(color=color_vector)

color_vector <- colorRampPalette(c("darkblue", "turquoise", "green", "yellow"))(n = 100)

color_vector
```

Find 69 with expression samples which also has genome sequencing data log2(1 + FPKM) (probably transformed already in from excel sheet)

Search for high and low expression characteristics over the samples

\* Divided each transcript into two groups (states) via k-means clustering

\* Only transcripts with at least 6 samples in each state should are continued with

\* having a state-average fold change larger than 3

-   avg(max(state_averages)) / average(min(state_averages)) \> 3
