---
title: "06 - Create heatmap analysis"
subtitle: "R for Bio Data Science"
format: html
author: Thomas Tams (s204540)
date: "2023-11-21"
editor: visual
---

## Load libraries

```{r}
library('tidyverse')
library("here")
source(here("R/99_proj_func.R"))
```

## Load data

```{r}
# Load data
data <- read_tsv(here("data/02_dat_clean.tsv"))
```

## Get expression data from all data

```{r}
expression_data <- data |> 
  select(1, 30:ncol(data))
```

## Log transform and transpose data to have transcript per row

```{r}
# Log2(1 + FPKM) transformation of expression value
# and transposing of the expression data
expr_log_trans <- expression_data |>
  group_by(sample_id) |>
  pivot_longer(-sample_id,
              names_to = "transcript_id",
              values_to = "expression_value") |>
  mutate(expression_value = log2(1 + expression_value)) |>
  pivot_wider(names_from = sample_id,
              values_from = expression_value) |>
  ungroup()
```

## Filter away transcripts with only zeros

```{r}
expr_log_trans_no_zero <- expr_log_trans |>
  filter(!if_all(starts_with("S"), ~ . == 0))
```

## Kmeans clustering on each transcript

For each individual transcript we split the expression values into two states, which should correspond to a high and low expression state.

```{r}
# Clustering function that clust
kmeans_cluster_row <- function(transcript_expr_data) {
  # Transpose tibble to fit into the kmeans function
  transposed_tibble <- transcript_expr_data |>
    pivot_longer(cols = everything(),
                 names_to = "sample_id")
  
  # Clustering
  cl <- transposed_tibble |>
    select(value) |>
    kmeans(centers=2)
  
  # Chain clustering states to each expression value
  cluster_tibble <- transposed_tibble |>
    mutate(cluster_int = cl$cluster) |>
    pivot_wider(names_from = sample_id,
                values_from = value)
  
  return(cluster_tibble)
}

# Perform kmeans clustering on each transcript
clust_data <- expr_log_trans_no_zero |>
  group_by(transcript_id) |>
  nest() |>
  mutate(data = map(.x = data, 
                    .f = ~kmeans_cluster_row(.x))) |>
  unnest(data) |>
  ungroup()

# Go to long format
clust_data_long <- clust_data |>
  pivot_longer(cols=-c(transcript_id, cluster_int),
               names_to = 'sample_id',
               values_to = 'expression_value') |>
  drop_na()
```

## Check list

```{r}
# Average expression fold change check
fold_change <- function(val_1, val_2){
  fold_change <- max(val_1, val_2) - min(val_1, val_2)
  return(fold_change)
}

avg_fold_change <- clust_data_long |>
  group_by(transcript_id, cluster_int) |>
  summarize(avg_expression = mean(expression_value)) |>
  ungroup(cluster_int) |>
  pivot_wider(names_from = cluster_int,
              values_from = avg_expression) |>
  rename(cluster_1_avg_expr = '1',
         cluster_2_avg_expr = '2') |>
  mutate(fold_change = map2(.x = cluster_1_avg_expr,
                            .y = cluster_2_avg_expr,
                            .f = ~fold_change(val_1 = .x,
                                              val_2 = .y))) |>
  ungroup()

# Save a list of transcript following 
# average expression fold change > 3
above_fold_change_threshold <- avg_fold_change |>
  filter(!fold_change == Inf) |>
  filter(fold_change > 3) |>
  pull(transcript_id)


# 6 expression values in each state check
count_clusters <- clust_data_long |>
  filter(transcript_id %in% above_fold_change_threshold) |>
  group_by(transcript_id) |>
  summarize(cluster1_n = sum(cluster_int == 1),
            cluster2_n = sum(cluster_int == 2)) |>
  ungroup()

# Save a list of transcript ids which have 6 or more
# expression values in each state
more_than_6_in_each_cluster <- count_clusters |>
  filter(cluster1_n >= 6 & cluster2_n >= 6) |>
  pull(transcript_id)

# Perform t-test and calculate false discovery rate q-values
clust_data_long_test <- clust_data_long |>
  filter(transcript_id %in% more_than_6_in_each_cluster) |>
  group_by(transcript_id) |>
  summarize(p_value = t.test(expression_value ~ cluster_int)$p.value) |>
  mutate(q_value = p.adjust(p_value, method = "BH"))

# Save a list of 
below_q_value <- clust_data_long_test |>
  filter(q_value < 0.01) |>
  pull(transcript_id)


clust_data_long |>
  filter(transcript_id %in% below_q_value) |>
  head(486) |>
  select(-sample_id) |>
  group_by(transcript_id) |>
  summarise(n=n())

clust_data_long |>
  filter(transcript_id %in% below_q_value) |>
  head(486) |>
  select(-sample_id) |>
  mutate(cluster_int = as.character(cluster_int)) |>
  group_by(transcript_id) |>
    summarise(
    p_value = fisher.test(table(expression_value, cluster_int))$p.value
  )


```

```{r}
selected_transcripts <- below_q_value
selected_transcripts |>
  as_tibble() |>
  write_tsv(here("data/below.tsv"))

if(!exists('selected_transcripts')){
  selected_transcripts <- 
    read_tsv(here("data/below.tsv"))
}

selected_transcripts <- selected_transcripts |>
  as_vector()
```

```{r}
# Sample data
set.seed(123)
gene_data <- data.frame(
  Group = rep(c("Group1", "Group2"), each = 10),
  Gene1 = sample(c(0, 1), 20, replace = TRUE),
  Gene2 = sample(c(0, 1), 20, replace = TRUE),
  Gene3 = sample(c(0, 1), 20, replace = TRUE),
  Gene4 = sample(c(0, 1), 20, replace = TRUE),
  Gene5 = sample(c(0, 1), 20, replace = TRUE),
  Gene6 = sample(c(0, 1), 20, replace = TRUE)
  # ... add more genes as needed
)

```

```{r}
# Reshape data for easier analysis
gene_data_long <- gene_data %>%
  pivot_longer(cols = -Group, names_to = "Gene", values_to = "Expression")

gene_data_long

# Perform Fisher's exact test for each gene
fisher_results <- gene_data_long |>
  group_by(Gene) |>
  summarise(
    p_value = fisher.test(table(Expression, Group))$p.value
  )

# Print the results
fisher_results

# Visualize the results
fisher_results %>%
  ggplot(aes(x = Gene, y = -log10(p_value))) +
  geom_bar(stat = "identity", fill = "skyblue") +
  labs(title = "Fisher's Exact Test for Gene Expression",
       x = "Gene",
       y = "-log10(p-value)")
```

## Save selected transcripts expression data to csv

```{r}
selected_transcript_expr <- expr_log_trans |>
  filter(transcript_id %in% selected_transcripts)

#selected_transcript_expr |>
#  write_tsv(here("data/06_selected_transcript_expr.tsv"))
```

## Load expression if already exists

```{r}
if(!exists('selected_transcript_expr')){
  selected_transcript_expr <- 
    read_tsv(here("data/06_selected_transcript_expr.tsv"))
}
```

```{r}
# check how many genes
selected_transcript_expr_split <- selected_transcript_expr |>
  separate(transcript_id, into = c("gene", "id"), sep = "_", extra="merge")

selected_transcript_expr_split |>
  group_by(gene) |>
  summarise(n = n()) |>
  arrange(desc(n))
```

```{r}

```

```{r}
expr_matrix <- as.matrix(selected_transcript_expr[, -1])
rownames(expr_matrix) <- selected_transcript_expr$transcript_id

color_vector <- colorRampPalette(c("darkblue", "turquoise", "green", "yellow"))(n = 100)

expr_matrix |>
  pheatmap::pheatmap(color=color_vector)



color_vector
```

Find 69 with expression samples which also has genome sequencing data log2(1 + FPKM) (probably transformed already in from excel sheet)

Search for high and low expression characteristics over the samples

-   Divided each transcript into two groups (states) via k-means clustering

<!-- -->

-   Only transcripts with at least 6 samples in each state should are continued with

<!-- -->

-   having a state-average fold change larger than 3

<!-- -->

-   avg(max(state_averages)) - average(min(state_averages)) \> 3
-   split values into cluster 1 and cluster 2 for each transcript
-   calculate test-t between values in each center for each transcript
    -   keep only transcript with q \> 0.01
-   Multiple gene transcript which meets requirements do either
    -   Select lowest q values to represent the gene (like in the paper)
    -   keep both ?
-   Do fisher exact test (threshold 1026) across transcripts and keep only transcripts with similar patterns in the two states with at least 6 other transcripts

```{r}
my_tibble <- tibble(
  Group = rep(c("A", "B"), each = 3),
  Value1 = c(10, 15, 20, 25, 30, 35),
  Value2 = c(5, 10, 15, 20, 25, 30)
)
```
